Squintability and Other Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2025-02-18 jso.qmd
update supplementary material
edit sec6

In addition, to assess the performance of JSO across various PP problems, two different data shapes, pipe and a sine wave, are considered in various dimensional spaces using six different PP indexes: `dcor2d_2`, `loess2d`, `MIC`, `TIC`,
`spline`, and `stringy`, under varied JSO hyper-parameters. This results
in a total of `r nrow(sim_df)` scenarios, comprising of
`r nrow(filter(sim_df, index == "holes"))` computed on the pipe data and
`r nrow(filter(sim_df, index != "holes"))` on the sine-wave data. This study does not consider dimensions lower than four, as exploring the search space in such cases is relatively straightforward. Similarly, extremely high-dimensional cases with complex data structures, where no method can reliably and consistenly uncover the underlying structure due to the curse of dimensionality, are also excluded, as they fall outside the scope of this study. A list of combinations of the data shape, dimensions and PP indexes that are considered in this study is provided in @tbl-smoothness-squintability. Again, JSO is run 50 times for each scenario to calculate the success rate for each hyper-parameter combination.

@@

To construct a relationship among success rate, smoothness,
squintability, and JSO hyper-parameters, a generalised linear model is
fitted using a binomial family and a logit link function. The data is
pre-processed by 1) scaling the JSO hyper-parameters by a factor of 10
for interpretation, 2) creating a new binary variable `long_time` to
indicate cases with an average run time over 20 seconds, and 3)
re-coding the success rate for the `stringy` index as 0, because none of
the 50 simulations correctly identified the sine-wave shape.

@@

The simulation result suggests that squintability and smoothness are likely dependent. Intuitively, a highly smooth search space exhibits gradual changes, making it easier to perceive the overall structure from a distance, thereby increasing the squint angle. @tbl-joint-output compare the model fits when squintability and smoothness are included separately. Given that the null deviance remains the same across models, the model with squintability yields a smaller residual deviance and lower AIC and BIC values, indicating a better fit. This suggests that squintability captures the relevant variation more effectively than smoothness.

>>>>

In addition, to assess the performance of JSO across various PP problems, two different data shapes, pipe and a sine wave, are considered in various dimensional spaces using six different PP indexes: `dcor2d_2`, `loess2d`, `MIC`, `TIC`,
`spline`, and `stringy`, under varied JSO hyper-parameters. An example of holes and pipes data with the syntax to generate them are provided in the supplementary material. This results
in a total of `r nrow(sim_df)` scenarios, comprising of
`r nrow(filter(sim_df, index == "holes"))` computed on the pipe data and
`r nrow(filter(sim_df, index != "holes"))` on the sine-wave data. This study does not consider dimensions lower than four, as exploring the search space in such cases is relatively straightforward. Similarly, extremely high-dimensional cases with complex data structures, where no method can reliably and consistenly uncover the underlying structure due to the curse of dimensionality, are also excluded, as they fall outside the scope of this study. A list of combinations of the data shape, dimensions and PP indexes that are considered in this study is provided in @tbl-smoothness-squintability. Again, JSO is run 50 times for each scenario to calculate the success rate for each hyper-parameter combination.

@@

To construct a relationship among success rate, smoothness,
squintability, and JSO hyper-parameters, a generalised linear model is
fitted using a binomial family and a logit link function. The data is
pre-processed by 1) scaling the JSO hyper-parameters by a factor of 10
for interpretation, 2) <!-- creating a new binary variable `long_time` to
indicate cases with an average run time over 20 seconds, and 3)-->
re-coding the success rate for the `stringy` index as 0, because none of
the 50 simulations correctly identified the sine-wave shape.

@@

The simulation result suggests that squintability and smoothness are likely dependent. Intuitively, a highly smooth search space exhibits gradual changes, making it easier to perceive the overall structure from a distance, thereby increasing the squint angle. @tbl-joint-output compare the model fits when squintability and smoothness are included separately. Given that the null deviance remains the same across models, the model with squintability yields a smaller residual deviance and lower AIC and BIC values, indicating a better fit. This suggests that squintability captures the relevant variation more effectively than smoothness. While it may seem counterintuitive at first that smoothness is not a significant factor, this outcome is reasonable in this context, as the swarm-based method used in this study, JSO, is designed to perform well even in non-smooth search spaces by leveraging population-based exploration.

### 2025-02-18 jso.tex
edit sec6

\begin{tcolorbox}[enhanced jigsaw, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, toprule=.15mm, coltitle=black, rightrule=.15mm, colback=white, toptitle=1mm, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, left=2mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, arc=.35mm, leftrule=.75mm, breakable, opacitybacktitle=0.6, titlerule=0mm]

@@

To construct a relationship among success rate, smoothness,
squintability, and JSO hyper-parameters, a generalised linear model is
fitted using a binomial family and a logit link function. The data is
pre-processed by 1) scaling the JSO hyper-parameters by a factor of 10
for interpretation, 2) creating a new binary variable
\texttt{long\_time} to indicate cases with an average run time over 20
seconds, and 3) re-coding the success rate for the \texttt{stringy}
index as 0, because none of the 50 simulations correctly identified the
sine-wave shape.

@@

The simulation result suggests that squintability and smoothness are
likely dependent. Intuitively, a highly smooth search space exhibits
gradual changes, making it easier to perceive the overall structure from
a distance, thereby increasing the squint angle.
Table~\ref{tbl-joint-output} compare the model fits when squintability
and smoothness are included separately. Given that the null deviance
remains the same across models, the model with squintability yields a
smaller residual deviance and lower AIC and BIC values, indicating a
better fit. This suggests that squintability captures the relevant
variation more effectively than smoothness.

>>>>

\begin{tcolorbox}[enhanced jigsaw, bottomtitle=1mm, left=2mm, toprule=.15mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, bottomrule=.15mm, toptitle=1mm, breakable, title={Algorithm: Jellyfish Optimizer Pseudo Code}, colback=white, titlerule=0mm, coltitle=black, arc=.35mm, rightrule=.15mm]

@@

To construct a relationship among success rate, smoothness,
squintability, and JSO hyper-parameters, a generalised linear model is
fitted using a binomial family and a logit link function. The data is
pre-processed by 1) scaling the JSO hyper-parameters by a factor of 10
for interpretation, 2) re-coding the success rate for the
\texttt{stringy} index as 0, because none of the 50 simulations
correctly identified the sine-wave shape.

@@

The simulation result suggests that squintability and smoothness are
likely dependent. Intuitively, a highly smooth search space exhibits
gradual changes, making it easier to perceive the overall structure from
a distance, thereby increasing the squint angle.
Table~\ref{tbl-joint-output} compare the model fits when squintability
and smoothness are included separately. Given that the null deviance
remains the same across models, the model with squintability yields a
smaller residual deviance and lower AIC and BIC values, indicating a
better fit. This suggests that squintability captures the relevant
variation more effectively than smoothness. While it may seem
counterintuitive at first that smoothness is not a significant factor,
this outcome is reasonable in this context, as the swarm-based method
used in this study, JSO, is designed to perform well even in non-smooth
search spaces by leveraging population-based exploration.
